---
title: "IS THERE GENDER BIAS?"
author: "Leroy Buliro"
date: "2/17/2020"
output: github_document
---

# IS THERE GENDER BIAS?
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1. INTRODUCTION

In this case study, we demystify the fact that male applicants are favoured more than their female counterparts when it comes to admissions. The data is provided by UC Berkeley.
We'll analyse the data by measuring the correlation between the two data sets used to claim gender bias then introduce a confounder to deny the hypothesis. Otherwise known as the Simpson's paradox.

## 2. METHODS AND ANALYSIS

## 2.1 Work Environment and Data Preparation

For this project, we'll use the following libraries

```{r message=FALSE}
if(!require(tidyverse)) 
  install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(dslabs)) 
  install.packages("dslabs", repos = "http://cran.us.r-project.org")
if(!require(broom)) 
  install.packages("broom", repos = "http://cran.us.r-project.org")
if(!require(knitr)) 
  install.packages("knitr", repos = "http://cran.us.r-project.org")
if(!require(rmarkdown)) 
  install.packages("rmarkdown", repos = "http://cran.us.r-project.org")
```

The required dataset is contained in the dslabs package which we just loaded. We proceed by loading the data using the code

```{r}
data(admissions)
```


## 2.2 Data Exploration
Here is the data

```{r}
admissions
```

It consists of

- The majors/departments include

```{r}
unique((admissions$major))
```

- A total number of **4526** applicants with **479** admitted

```{r echo=FALSE}
totals <- admissions %>% group_by(gender) %>% 
  summarize(admitted=sum(admitted), applicants=sum(applicants))
totals %>% summarize(total_admitted=sum(admitted), 
                     total_applicants=sum(applicants))
```

- Among men and women

```{r echo=FALSE}
admissions %>% group_by(gender) %>% 
  summarize(admitted=sum(admitted), applicants=sum(applicants))
```

Percentage wise **44.5%** of total men and **30.3%** of total women had their applications approved

```{r}
admissions %>% group_by(gender) %>% 
  summarize(percentage = 
              round(sum(admitted*applicants)/sum(applicants),1))
```


## 2.3 Data Analysis and Visualization

We begin by testing whether gender and admission are independent

```{r}
admissions %>% group_by(gender) %>% 
  summarize(total_admitted = round(sum(admitted / 100 * applicants)), 
            not_admitted = sum(applicants) - sum(total_admitted)) %>% 
  select(-gender) %>% 
  do(tidy(chisq.test(.)))
```

A statistical test, the chi-squared test, clearly rejects the hypothesis that gender and admissions are independent as the p value (**1.06e-21**) is very small.

Following the findings above, we now evaluate the difference in admissions by gender

```{r}
admissions %>% select(major, gender, admitted) %>% 
  spread(gender, admitted) %>%
  mutate(women_minus_men = women - men)
```

Four majors favor women over men. These are **A, B, D and F**.

The paradox is that analyzing the totals suggest a dependence between admissions and gender. But when the data is grouped by major, this dependence seems to disappear. Clearly, Z (major) is associated with Y (females), because the more selective a major, the lower the probability that someone enters that major.

Is major selectivity associated with gender? Let's plot total percent admitted to major versus percent women applicants

```{r echo=FALSE}
admissions %>% 
  group_by(major) %>% 
  summarize(major_selectivity = sum(admitted * applicants) / sum(applicants),
            percent_women_applicants = sum(applicants * (gender=="women")) /
              sum(applicants) * 100) %>%
  ggplot(aes(major_selectivity, percent_women_applicants, label = major)) +
  geom_text()
```

The plot suggests that women were much more likely to apply to the two hard majors. Gender and major selectivity are confounded.
Major E is much harder to enter than major B. And over 60% of applicants to major E were women, while less than 30% of the applicants of major B were women.

```{r echo=FALSE}
admissions %>%
  mutate(yes = round(admitted/100*applicants), no = applicants - yes) %>%
  select(-applicants, -admitted) %>%
  gather(admission, number_of_students, -c("major", "gender")) %>%
  ggplot(aes(gender, number_of_students, fill = admission, ggtitle("Number of Applicants Admitted and Not"))) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(. ~ major)
```

```{r echo=FALSE}
admissions %>% 
  mutate(percent_admitted = admitted * applicants/sum(applicants)) %>%
  ggplot(aes(gender, y = percent_admitted, fill = major)) +
  geom_bar(stat = "identity", position = "stack")
```

This breakdown lets us see that the majority of accepted men came from two majors, A and B. It also lets us see that few women apply to these two easy majors.

Now let us compare the number of applicants per gender

```{r echo=FALSE}
admissions %>% 
  ggplot(aes(major, admitted, col = gender, size = applicants)) + geom_point()
```

The size of the dot represents the number of applicants, and explains the paradox. We see large red dots and small blue dots for the easiest majors, A and B.

If we first stratify by major, compute the difference, and then average, we find that the percent difference is actually quite small.

```{r}
admissions %>%  group_by(gender) %>% summarize(average = mean(admitted))
```


## 4. CONCLUSION

In our first analysis we see that **44.5%** of men and **30.3%** of women got accepted from the total number of applicants. But on further analysis, if we first stratify by major, compute the difference, and then average, we find that the percent difference is actually quite small with **38.2%** men and **41.7%** women. We can then conclude that there is no gende bias.